---
title: "Plotting trees with ggtree"
author: Raphael Sieber
affiliation: Statens Serum Institut
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

This is an example script for plotting a phylogenetic tree with cluster highlights and metadata annotations. The analysis involves loading and rooting the tree, preparing metadata, and visualizing the tree in both circular and rectangular layouts.

# Load Required Packages

```{r load-packages}
# Load the necessary libraries
library(tidyverse)
library(phytools)
library(ggtree)
```

# Load and Prepare the Phylogenetic Tree

## Load the Tree

```{r load-tree}
# Load the phylogenetic tree from a file
# Replace with the correct path to your tree file
tree <- read.tree(file.path("data", "core_stripped.filtered_polymorphic_sites.fasta.treefile"))

# Plot the initial unrooted tree
plot(tree)
```

## Root the Tree

```{r root-tree}
# Root the tree at a specific node defined by a group of strains
rooted <- reroot(tree, node.number = getMRCA(phy = tree, c(
  "SRR27240808", "SRR27240815", "SRR27240809", "SRR27240826",
  "SRR27240827", "SRR27240813", "SRR27240814", "SRR27240817",
  "SRR27240822", "SRR27240818", "SRR27240819"
)), position = 0.0015)

# Plot the newly rooted tree
plot(rooted)
```

# Load and Prepare Metadata

## Load Metadata

```{r load-metadata}
# Load metadata for the strains
# Replace with the correct path to your metadata file
meta <- read.delim(file.path("data", "metadata.tsv"), sep='\t')
```

## Group and Process Metadata

```{r process-metadata}
# Group the metadata by clusters and find the MRCA for each cluster
clusters <- meta %>%
  group_by(SNPCluster) %>%
  mutate(cl_mrca = getMRCA(rooted, strain)) %>%
  filter(!is.na(SNPCluster))

# Display the clustered metadata
clusters
```


# Define colors and functions
## Define cluster colors

```{r define-colors}
cl_cols <- c(`ST1#1`="#1F78C8", `ST1#10_1`="#33a02c", `ST1#10_2`="#FFD700",
             `ST1#11`="#ff7f00", `ST1#9`="#6A33C2")
```

## Define helper Function
This function puts the cluster highlights in the background, so they don't cover for branches and labels.  

```{r helper-functions}
# Helper function to bring highlight layers to the back
highlight_to_back <- function(p){
  pl <- p
  hl <- sapply(pl$layers, function(x) grepl("Hilight", class(x$geom)[1]))
  pl$layers <- c(pl$layers[hl], pl$layers[!hl])
  return(pl)
}
```

# Circular Phylogenetic Tree

## Create Circular Tree

```{r circular-tree-creation}
# Create a circular tree with ggtree
p <- ggtree(rooted, layout="c", right = TRUE, lwd=0.4, show.legend = FALSE, open.angle = 10)

# Rotate the tree slightly for better visualization
q <- rotate_tree(p, angle = 10) %>%
  groupClade(clusters$cl_mrca, 'cl') +
  geom_tippoint(size=0.6, show.legend = FALSE) +
  geom_cladelab(
    data = clusters,
    mapping=aes(node=cl_mrca, label=SNPCluster, color=SNPCluster),
    fontsize = 4, align = TRUE, show.legend = FALSE, barcolour=NA, hjust = 0.5, offset = 0.0015
  ) +
  scale_color_manual(values = cl_cols) +
  scale_fill_manual(values = cl_cols) +
  xlim(-0.002, 0.01) +
  geom_hilight(
    data = clusters, mapping=aes(node=cl_mrca, fill=SNPCluster),
    extendto= max(p$data$x)
  ) +
  geom_tiplab(align = TRUE, offset = 0.002, size = 3) +
  geom_treescale(width = 0.001, y = Ntip(rooted), x = 0.0001) +
  geom_rootedge(rootedge = 0.001)
```

## Adjust Highlights

```{r move-highlights-to-back}
# Adjust the order of highlighted layers
r <- q %>% highlight_to_back()
```

## Plot circular tree

```{r plot-circular-tree}
r
```

## Save circular tree

```{r save-circular-tree}
# Save the circular tree as a PDF
ggsave(r, width = 8, height = 8, device = cairo_pdf,
       file=file.path('results', paste0("Tree_circ_w_clusters_", Sys.Date(), ".pdf")),
       family = "sans", bg=NA)
```

# Rectangular Phylogenetic Tree

## Setup for Metadata Annotations

```{r rectangular-tree-setup}
# Table setup for metadata annotations
tab_cols <- c("SamplingDate" = "date", "Region" = "region")
table_df <- data.frame(
  cbind(tab_cols, tab_cols, 1), stringsAsFactors = FALSE
)
colnames(table_df) <- c("name", "ring_label", "rel_width")
table_df$ring_label <- names(tab_cols)

# Plotting parameters for metadata annotations
tabv <- data.frame(
  offs_s = 0.0033,                       # The distance between the tree and the first column
  ann_yjust = 1.0, ann_xjust = 0.000001, # adjust the column header position
  ann_vjust = 0.3, ann_hjust = 0,        # adjust the column header anchor
  ann_size = 2,                          # Text size for column header
  xdiff = 0.0013,                        # Space from one column to the next
  size = 2                               # Text size for column values
)

# Calculate widths and positions for metadata columns
table_df$width <- as.numeric(table_df$rel_width) * tabv$xdiff
table_df$x_start <- cumsum(c(0, table_df$width[1:(nrow(table_df)-1)]))
```

## Create and Annotate Rectangular Tree

```{r rectangular-tree-creation}
# Create a rectangular tree with metadata columns
p <- ggtree(rooted, layout="rectangular", right = FALSE, lwd=0.3) %<+% meta
p <- p +
  geom_tiplab(size=2, offset = 0.002, align = TRUE, linetype = NA) +
  geom_tippoint(color="black", cex=0.5) +
  xlim(-0.0001, 0.01) +
  geom_treescale(width = 0.0001)

# Add cluster labels
for(i in 1:nrow(clusters))
  p <- p + geom_cladelabel(
    node = clusters$cl_mrca[i], label= clusters$SNPCluster[i],
    barsize=0.5, color="black", offset.text=0, extend = 0.4, align = TRUE, fontsize = 4, offset = 0
  )

# Add cluster highlights
q <- p + geom_hilight(
  data = clusters, mapping=aes(node=cl_mrca, fill=SNPCluster),
  extendto= max(p$data$x)
) + scale_fill_manual("Lineage", values = cl_cols) +
  theme(legend.position = 'None')
```

## Add Metadata Columns

```{r add-metadata-to-tree}
# Add metadata columns
for(j in 1:nrow(table_df)) {
  q <- q + geom_tiplab(
    aes(label=.data[[table_df[j, "name"]]]), size=tabv$size, geom="text",
    align=TRUE, linesize=0.05, linetype=NA, show.legend = FALSE,
    offset=tabv$offs_s + table_df[j, "x_start"]
  ) +
    # Also add column headers
    annotate(
      geom="text", x=max(p$data$x) + tabv$offs_s + table_df[j, "x_start"] + tabv$ann_xjust,
      y = max(p$data$y) + tabv$ann_yjust, label=table_df[j, "ring_label"],
      size=tabv$ann_size, angle=0, vjust=tabv$ann_vjust, hjust=tabv$ann_hjust
    )
}
```

## Adjust Highlights

```{r move-highlights-to-back2}
# Adjust the order of highlighted layers
r <- q %>% highlight_to_back()
```

## Plot the tree

```{r plot-rectangular-tree}
r
```

## Save the tree as pdf

```{r save-rectangular-tree}
ggsave(r, width = 12, height = 42, device = cairo_pdf,
       file=file.path('results', paste0("Tree_rect_w_clusters_", Sys.Date(), ".pdf")),
       family = "sans", bg=NA)
```
