---
title: "root-to-tip-analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Perform a root-to-tip analysis using R

```{r load packages}
library(tidyverse)
library(phytools)
```

```{r load the tree}
phy <- read.tree(file.path("data", "core_stripped.filtered_polymorphic_sites.fasta.treefile"))
plot(phy)
rooted <- reroot(phy, node.number = getMRCA(phy = phy, c(
  "SRR27240808",
  "SRR27240815",
  "SRR27240809",
  "SRR27240826",
  "SRR27240827",
  "SRR27240813",
  "SRR27240814",
  "SRR27240817",
  "SRR27240822",
  "SRR27240818",
  "SRR27240819"
)),position = 0.0015)
plot(rooted)
add.scale.bar()
```

## Export the rooted tree

```{r export-rooted-tree}
write.tree(rooted, file = file.path('results', 'core_stripped.filtered_polymorphic_sites.fasta.treefile.rooted.nwk'))
```

## Load metadata

```{r load-metadata}
meta <- read.delim(file.path("data", "metadata.tsv"), sep='\t') %>%
  mutate(
    date = paste(date, '01', sep='-')
  )
meta
```

## Get tip dates

```{r get-tip-dates}
tip.dates <- as.Date(meta[match(phy$tip.label, meta$strain), ]$date, format="%Y-%m-%d")
```

## Get divergence from root

```{r get-divergence}
div <- node.depth.edgelength(rooted)[1:Ntip(rooted)]
```

## Get the correlation between divergence and date

```{r lm}
m <- lm(div ~ tip.dates)
```

## Plot divergence vs. sample dates

```{r plot-data}
plot(tip.dates, div, xlim=c(as.Date("2015-01-01"), max(tip.dates, na.rm = T)),
     ylim=c(0, max(div)))
abline(m)
```

## Get the correlation coefficient R^2^

```{r get-r-squared}
summary(m)$r.squared
```
>**Note:** This is a very bad correlation and a negative slope. To properly infer phylogenetic time scales, a high R^2^ and a positive slope is required.  
